### 2023-02-03 paper-thesis/chapters/01-introduction.tex
Start reworking spoofax chapter

Before explaining these contributions, we provide background information on Spoofax and Statix (Chapter \ref{chap:bg-dp}) and the Calculus of Constructions (Chapter \ref{chap:bg-spoofax}).

>>>>

\noindent Before explaining these contributions, we provide background information on Spoofax and Statix (Chapter \ref{chap:bg-dp}) and the Calculus of Constructions (Chapter \ref{chap:bg-spoofax}).

The source code for this thesis can be found online: \url{https://github.com/JonathanBrouwer/master-thesis}

### 2023-02-03 paper-thesis/chapters/02-background-spoofax.tex
A substitution based approach
Start reworking spoofax chapter

This chapter will explain the concepts behind the Spoofax Language Workbench. If you are already familiar with Spoofax, you can skip this chapter.

\section{Spoofax}

The Spoofax Language Workbench~\cite{spoofax} is a platform to develop domain-specific and
general-purpose programming languages. Each aspect of the language is defined in one of three metalanguages, each with their own purpose. From these languages Spoofax generates a parser, type checker and other useful tools.

The metalanguage SDF3 is used to define the syntax of the language~\cite{sdf3}, which is then parsed using Spoofax's SGLR parser~\cite{sdf3_parser}. The parser produces a abstract syntax tree in the ATerm format~\cite{aterm}. A pretty printer is also generated based on this syntax definition.

Next, the metalanguage Statix~\cite{scopes_as_types} is used to define the static semantics of the language. It uses scope graphs to represent binding and typing information of the language. How this works is discussed in section \ref{sec:statix}.

Finally, Stratego~\cite{stratego} is used to define program transformations, using rewrite rules. This can be used for modifications to the AST, but also to build transformations to a different language (such as compiling to assembly).

\section{\label{sec:statix}Statix and Scope graphs}

\textit{Note: This section is adapted from the Explanation in "Composable Type System Specification
	using Heterogeneous Scope Graphs".~\cite[sect. 4.1.2]{Zwaan21}}

In Statix, scope graphs are used to represent a programs name and type information~\cite{scopes_as_types}. A scope graph is a graph that consists of the following components:

@@

Information can be retrieved from scope graphs using \textit{queries}. A query from a scope traverses the scope graph to find datums that match certain conditions. The result of a query is a list of (path, datum) tuples. A query has several parameters:
\begin{itemize}
	\item The relation to query. Only datums declared under this relation will be returned by the query.
	\item Path well-formedness condition: a regular expression of labels that specifies which paths are well-formed. Only datums that are reached through a path whose edge labels are in the language described by the regular expression are included in the query result.
	\item A match predicate, taking a single datum as input. Only datums that satisfy this predicate will be included in the query result. Its default value is \verb|true|, meaning that all datums under the specified relation that satisfy the path well-formedness condition are returned.
	\item Result comparison parameters:
	\begin{itemize}
		\item A strict partial order on paths, described by less-than relations on labels. This relation defines a prefix-order on paths.
		\item A `shadow' predicate, which takes two datums as input. Its default value is \verb|false|, meaning that no shadowing is performed.
	\end{itemize}
	When the result set contains multiple datums, and it holds for a datum $d$ that there exists another datum $d'$ which path is strictly smaller that $d$, and the shadow predicate holds, then $d$ is removed from the result set.
\end{itemize}

>>>>

This chapter will explain the concepts behind the Spoofax Language Workbench. Readers already familiar with Spoofax are recommended to only skim through this chapter.

The Spoofax Language Workbench~\cite{spoofax} is a platform to develop domain-specific and
general-purpose programming languages. Each aspect of the language is defined in one the meta-languages, each with their own purpose. From these specifications Spoofax generates a parser, type checker and other useful tools. The meta-languages that are used in this thesis are discussed in the sections below:

\section{SDF3}

The meta-language SDF3 is used to define the syntax of the language~\cite{sdf3}, which is then parsed using Spoofax's SGLR parser~\cite{sdf3_parser}. The parser produces a abstract syntax tree in the ATerm format~\cite{aterm}. A pretty printer is also generated based on this syntax definition.

\paragraph{Syntax definitions}
The syntax of

\paragraph{ATerm format}

\section[Statix and Scope graphs]{\label{sec:statix}Statix and Scope graphs\protect\footnote{Note: This section is adapted from "Composable Type System Specification using Heterogeneous Scope Graphs".~\cite[sect. 4.1.2]{Zwaan21}}}

In Statix, scope graphs are used to represent a programs name and type information~\cite{scopes_as_types}~\cite{nameres}. A scope graph is a graph that consists of the following components:

@@

Information can be retrieved from scope graphs using \textit{queries}. A query from a scope traverses the scope graph to find datums that match certain conditions. The result of a query is a list of (path, datum) tuples. A query has several parameters:
\begin{itemize}
	\item The relation to query. Only datums declared under this relation will be returned by the query.
	\item Path well-formedness condition: a regular expression of labels that specifies which paths are well-formed. Only datums that are reached through a path whose edge labels are in the language described by the regular expression are included in the query result.
	\item A match predicate, taking a single datum as input. Only datums that satisfy this predicate will be included in the query result. Its default value is \verb|true|, meaning that all datums under the specified relation that satisfy the path well-formedness condition are returned.
\end{itemize}
